---
title: "Report about conserved transcriptional signatures and evolutionary relationships of germ/stem cells in the PIWI-piRNA pathway"
author: "Constantinos Yeles (Konstantinos Geles)"
date: "Last update : `r format(Sys.Date(), '%a %b %d %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    theme: paper 
    keep_md: true
  pdf_document:
    toc: yes
    toc_depth: 5
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Conserved transcriptional signatures and evolutionary relationships of germ/stem cells
## The PIWI-piRNA pathway expression in germ/stem cells

Using the [Expression Atlas Bioconductor Package](https://www.bioconductor.org/packages/release/bioc/html/ExpressionAtlas.html) we will 
search about the expression profiles of genes that are involved in the [piRNA biogenesis](https://reactome.org/PathwayBrowser/#/R-HSA-75944&SEL=R-HSA-163316&PATH=R-HSA-74160) in various organisms and conditions.

### Materials and Methods
The workflow has been primarily carried out on a Linux server, 
but it can be used easily on a Windows or Mac OS machine with plenty of RAM.

The workflow utilizes _[R](https://www.r-project.org/)_ scripting for various operations.
For the application of the workflow, the following tools/ workflows have been used:

*  _[Rstudio](https://rstudio.com/)_ for R scripting,
*  _**[Expression Atlas](https://www.bioconductor.org/packages/release/bioc/html/ExpressionAtlas.html)**_ for getting the datasets,
*  _[Creating a network of human gene homology with R and D3](https://shiring.github.io/genome/2016/12/11/homologous_genes_post)_ for the identification of gene IDs and homologous genes between species with the help of  **[biomaRt](https://bioconductor.org/packages/release/bioc/html/biomaRt.html)** and **[AnnotationDbi](https://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html)**

## Workflow  

### Install and load libraries  

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("ExpressionAtlas", "biomaRt", "AnnotationDbi", 
                       "EnsDb.Hsapiens.v86", "tidybulk", "tidySummarizedExperiment"))

install.packages(c("tidyverse", "magrittr"))

suppressPackageStartupMessages({
  library('ExpressionAtlas')
  library('biomaRt')
  library('AnnotationDbi')
  library('EnsDb.Hsapiens.v86')
  library('tidybulk')
  library('tidySummarizedExperiment')
  library('tidyverse')
  library('magrittr')
})
```

### Search which datasets in Expression Atlas have "stem" and "germ" terms
```{r}
atlasRes_stem <- searchAtlasExperiments( properties = "stem")
atlasRes_germ <- searchAtlasExperiments( properties = "germ")
```

Search if there are identical datasets in both sets 
```{r}
atlasRes_germ %>% as_tibble %>% filter(is_in(Accession, atlasRes_stem$Accession))
```

join the two sets
```{r}
atlasRes <- as_tibble(atlasRes_stem) %>% full_join(as_tibble(atlasRes_germ))

rm(atlasRes_stem, atlasRes_germ)
```

Search which species are in the dataset
```{r}
atlasRes %>% dplyr::count(Species, sort = TRUE) 
```

We will remove plant species from the dataset and microarray experiments
```{r}
atlasRes <- atlasRes %>% dplyr::filter(is_in(Species, 
                                          c("Mus musculus",
                                            "Drosophila melanogaster",
                                            "Rattus norvegicus",
                                            "Caenorhabditis elegans")),
                                str_detect(Type, "RNA-seq"))
```

### Get the AtlasData summaries
```{r}
rnaseqExps <- getAtlasData(atlasRes$Accession)
```

### scale the counts using tidybulk
```{r}
datasets_atlass_scaled <- map(
    names(rnaseqExps)[1:2] %>% 
         purrr::set_names(),
    ~ rnaseqExps %>% 
        extract2(.x) %>% 
        extract2("rnaseq") %>%
        identify_abundant(factor_of_interest = AtlasAssayGroup) %>% 
        scale_abundance()
)

rm(atlasRes, rnaseqExps)
```

### Following _**Shirin's playgRound**_ instructions on getting gene identifiers
*with small modifications 
#### Use biomaRt to access the datasets of species we prefere
```{r}
ensembl = useMart("ensembl")

datasets <- listDatasets(ensembl) %>% 
    dplyr::filter(str_detect(dataset, 
                  c("mmusculus|rnorvegicus|melanogaster|celegans"))) 


human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
specieslist  <-  datasets$dataset
```

#### Make an object for each species of interest from the Ensembl database
```{r}
for (i in seq_along(specieslist)) {
    print(specieslist[i])
    ensembl <- datasets[i, 1]
    assign(paste0(ensembl), useMart("ensembl", dataset = paste0(ensembl)))
}
```

#### Identify human gene homologs for the Protein coding genes in piRNA biogenesis pathway
```{r}
# get all Ensembl gene IDs
human_EnsDb <- keys(EnsDb.Hsapiens.v86, keytype = "GENEID")

# IDs of piRNA biogenesis pathway genes
PIWI_pirna_biogenesis <- c("P24928", "P62487", "P52435", "P36954", "O15514",
                           "P61218", "P52434", "P62875", "P19387", 'P53803',	
                           "P19388", "P30876", 'Q5T8I9', "Q7Z3Z4", "Q96JY0",
                           "Q9Y2W6", "Q8NDG6", "Q8TC59", "Q587J7", "Q8WWH4",
                           "Q9NQI0", "Q9BXT6", "Q9BXT4", "Q96J94", "O60522", 
                           "Q8N2A8", "P07900", "O75344", "P10243")

# select on the above identifiers
gene_symbols <- AnnotationDbi::select(EnsDb.Hsapiens.v86, 
                       keys = human_EnsDb, 
                       columns = c("SYMBOL", "UNIPROTID", "GENEBIOTYPE")) %>% 
    dplyr::filter(UNIPROTID %in% PIWI_pirna_biogenesis)
```

Get dataframe of human genes and their homologs 
```{r}
for (species in specieslist) {
    print(species)
    assign(paste0("homologs_human_", species), getLDS(attributes = c("ensembl_gene_id", "chromosome_name"), 
                                                      filters = "ensembl_gene_id", 
                                                      values = gene_symbols$GENEID, 
                                                      mart = human, 
                                                      attributesL = c("ensembl_gene_id", 
                                                                      "chromosome_name", 
                                                                      "external_gene_name"), 
                                                      martL = get(species)))
}

homologs_human_celegans_gene_ensembl <-  homologs_human_celegans_gene_ensembl %>% 
    mutate(organism_mart = "celegans") %>% 
    as_tibble() %>% 
    mutate(Chromosome.scaffold.name.1 = as.character(Chromosome.scaffold.name.1))
    
homologs_human_dmelanogaster_gene_ensembl <- 
    homologs_human_dmelanogaster_gene_ensembl %>% 
    mutate(organism_mart = "dmelanogaster") %>% 
    as_tibble() %>% 
    mutate(Chromosome.scaffold.name.1 = as.character(Chromosome.scaffold.name.1))

homologs_human_mmusculus_gene_ensembl <- 
    homologs_human_mmusculus_gene_ensembl %>%
    mutate(organism_mart = "mmusculus") %>% 
    as_tibble()%>% 
    mutate(Chromosome.scaffold.name.1 = as.character(Chromosome.scaffold.name.1))

homologs_human_rnorvegicus_gene_ensembl <- 
    homologs_human_rnorvegicus_gene_ensembl %>% 
    mutate(organism_mart = "rnorvegicus") %>% 
    as_tibble()%>% 
    mutate(Chromosome.scaffold.name.1 = as.character(Chromosome.scaffold.name.1))

piwi_genes <- dplyr::bind_rows(homologs_human_celegans_gene_ensembl,
                                homologs_human_dmelanogaster_gene_ensembl,
                                homologs_human_mmusculus_gene_ensembl,
                                homologs_human_rnorvegicus_gene_ensembl)
```

Keep only the piRNA biogenesis genes from the datasets
```{r}
 piwi_genes_dat <- map(datasets_atlass_scaled,
    ~filter( .x, is_in(.feature, piwi_genes$Gene.stable.ID.1)) %>% 
    left_join(distinct(piwi_genes, Gene.stable.ID.1, .keep_all = "TRUE"),
              by = c(.feature = "Gene.stable.ID.1" ))
    )

#########################

atlasRes %>% 
    names %>%  
    set_names() %>% 
    map( ~dplyr::count(as_tibble(atlasRes), .data[[.x]], sort = TRUE))


```



